prolfqua&colon; A Comprehensive R-package for Proteomics Differential Expression Analysis

### 2023-01-09 inst/MyArticle/SIB.Rmd
rename ContrastSimpleImpute to ContrastMissing #28

## specify differences among groups
contrasts <- c(
    "dilution_(9/7.5)_1.2" = "dilution.e - dilution.d",
    "dilution_(7.5/6)_1.25" = "dilution.d - dilution.c" )
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- 
    prolfqua::strategy_lm(lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)
## compute differences from linear model
## and moderate variance
contr <- prolfqua::Contrasts$new(models, contrasts) |> 
    prolfqua::ContrastsModerated$new()
## compute differences using imputation
conI <- 
    prolfqua::ContrastsSimpleImpute$new(lfqp, contrasts) |> 
    prolfqua::ContrastsModerated$new()
## merge contrasts, to obtain differences for all proteins
contrasts <- prolfqua::addContrastResults(
    prefer = contr, add = conI)
## plot panel E and F
pl <- contrasts$merged$get_Plotter() 
panelE <- pl$histogram()$p.value + ggplot2::labs(tag = "E")                                                        
panelF <- pl$volcano()$FDR +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(tag = "F")
gridExtra::grid.arrange(panelA, panelB,
                        panelC, panelD,
                        panelE, panelF, ncol = 2)
```

@@

```{r describeBenchmarkedModels}
xMM <- data.frame(
    msstats = c("MSstats", "preprocess with default parameters",  "", "evidence.txt"),
    proDA = c("proDA", "probabilistic dropout model",  "protein", "peptide.txt"),
    imputation = c("prot_imputation", "ContrastsSimpleImpute, ContrastsModerated", "protein",  "peptide.txt"),
    lmmed = c("prot_med_lm_moderated", "strategy_lm, Contrasts, ContrastsModerated",  "protein", "peptide.txt"),
    prot_merged = c("prot_merged", "addContrastResults(prefer = prot_med_lm_moderated , add = prot_imputation) *",  "protein", "peptide.txt"),
    mixed = c("prot_mixed_effect_moderated", "strategy_lmer, Contrasts, ContrastsModerated", "peptide", "peptide.txt" ),
    ropeca = c("ropeca", "strategy_lm, Contrasts, ContrastsModerated, ContrastsROPECA", "peptide", "peptide.txt" )
)

>>>>

## specify differences among groups
contrasts <- c(
    "dilution_(9/7.5)_1.2" = "dilution.e - dilution.d",
    "dilution_(7.5/6)_1.25" = "dilution.d - dilution.c" )
## fit model
lmmodel <- paste(lfqp$intensity_column()," ~ dilution.")
modelFunction <- 
    prolfqua::strategy_lm(lmmodel, model_name = "lm")
models <- prolfqua::build_model(lfqp, modelFunction)
## compute differences from linear model
## and moderate variance
contr <- prolfqua::Contrasts$new(models, contrasts) |> 
    prolfqua::ContrastsModerated$new()
## compute differences using imputation
conI <- 
    prolfqua::ContrastsMissing$new(lfqp, contrasts) |> 
    prolfqua::ContrastsModerated$new()
## merge contrasts, to obtain differences for all proteins
contrasts <- prolfqua::addContrastResults(
    prefer = contr, add = conI)
## plot panel E and F
pl <- contrasts$merged$get_Plotter() 
panelE <- pl$histogram()$p.value + ggplot2::labs(tag = "E")                                                        
panelF <- pl$volcano()$FDR +
    ggplot2::theme(legend.position = "bottom") +
    ggplot2::labs(tag = "F")
gridExtra::grid.arrange(panelA, panelB,
                        panelC, panelD,
                        panelE, panelF, ncol = 2)
```

@@

```{r describeBenchmarkedModels}
xMM <- data.frame(
    msstats = c("MSstats", "preprocess with default parameters",  "", "evidence.txt"),
    proDA = c("proDA", "probabilistic dropout model",  "protein", "peptide.txt"),
    imputation = c("prot_imputation", "ContrastsMissing, ContrastsModerated", "protein",  "peptide.txt"),
    lmmed = c("prot_med_lm_moderated", "strategy_lm, Contrasts, ContrastsModerated",  "protein", "peptide.txt"),
    prot_merged = c("prot_merged", "addContrastResults(prefer = prot_med_lm_moderated , add = prot_imputation) *",  "protein", "peptide.txt"),
    mixed = c("prot_mixed_effect_moderated", "strategy_lmer, Contrasts, ContrastsModerated", "peptide", "peptide.txt" ),
    ropeca = c("ropeca", "strategy_lm, Contrasts, ContrastsModerated, ContrastsROPECA", "peptide", "peptide.txt" )
)

### 2023-01-09 inst/MyArticle/analysingDataStartingfromProteinGroups.Rmd
rename ContrastSimpleImpute to ContrastMissing #28

```{r}
lfqTrans$config$table$factorDepth <- 2
# ContrastsSimpleImpute$debug("get_contrasts")
contrSimple <- prolfqua::ContrastsSimpleImpute$new(lfqdata = lfqTrans,
                                                   Contrasts)
contrdfSimple <- contrSimple$get_contrasts()
# na.omit(contrdfSimple)
pl <- contrSimple$get_Plotter()
pl$histogram_diff()
pl$volcano()
```

>>>>

```{r}
lfqTrans$config$table$factorDepth <- 2
# ContrastsMissing$debug("get_contrasts")
contrSimple <- prolfqua::ContrastsMissing$new(lfqdata = lfqTrans,
                                                   Contrasts)
contrdfSimple <- contrSimple$get_contrasts()
# na.omit(contrdfSimple)
pl <- contrSimple$get_Plotter()
pl$histogram_diff()
pl$volcano()
```

### 2023-01-09 inst/MyArticle/paper.Rmd
rename ContrastSimpleImpute to ContrastMissing #28

*R*'s formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with *prolfqua* in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in *prolfqua* (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform DEA given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsSimpleImpute`). Further implementations of the interface encapsulate and integrate DEA results of external tools such as *proDA* or of *SAINTexpress* [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

```{r prolfquaModels}
xM <- data.frame( 
    lm = c("strategy_lm, Contrasts", "linear modelling of peptide or protein abundances and group difference estimation"),
    lmer = c("strategy_lmer, Contrasts", "mixed effect modelling of peptide or protein abundances and group differences estimation" ),
    gmi = c("ContrastsSimpleImpute" , "group difference estimation when no observations in one of the groups" ),
    ROPECA = c("ContrastsROPECA", "estimating group differences for proteins by summarizing peptide differences"),
    limma = c( "ContrastsModerated", "empirical Bayes variance shrinkage for group difference estimates (limma)"),
    pic = c("runSaint, ContrastsSaintExpress", "protein interaction scoring (SAINTExpress)"),
    proDA = c( "strategy_proDA*, ContrastsProDA", "adapter to the probabilistic dropout model implemented in proDA"), check.names =  FALSE)
xM <- data.frame(t(xM))

@@

The following code example illustrates how we compute differences among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsSimpleImpute` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `merge_contrasts_results` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:figure2). Panel A shows the distribution of the $p$-values, Panel B shows the volcano plot for each comparison, and Panel C shows a Bland-Altman plot reporting the difference between groups as a function of the rank of the protein abundance.

@@

## compute contrasts using imputation imputation
conI <- prolfqua::ContrastsSimpleImpute$new( lfqpro, contrasts, modelName = "imp")

@@

allBenchmarks <- allBenchmarks[c("benchmark_imputation","benchmark_ProtModerated",  "benchmark_mixedModerated", "benchmark_ropeca","benchmark_merged","benchmark_mssstats","benchmark_proDA", "benchmark_msqrob")]

@@

```{r describeBenchmarkedModels, results='asis' }
xMM <- data.frame(
    msstats = c("MSstats", "preprocess with default parameters",  "", "evidence.txt"),
    msqrob = c("msqrob2", "hurdle msqrob model", "protein and peptide", "peptide.txt"),
    proDA = c("proDA", "probabilistic dropout model",  "protein", "peptide.txt"),
    imputation = c("prolfqua_estimate", "ContrastsSimpleImpute, ContrastsModerated", "protein",  "peptide.txt"),
    lmmed = c("prolfqua_lm_mod", "strategy_lm, Contrasts, ContrastsModerated",  "protein", "peptide.txt"),
    prot_merged = c("prolfqua_merged", "addContrastResults( prefer = proflqua_lm_mod , add = prolfqua_estimate)HACKIT ",  "protein", "peptide.txt"),
    mixed = c("prolfqua_mix_eff_mod", "strategy_lmer, Contrasts, ContrastsModerated", "peptide", "peptide.txt" ),
    ropeca = c("prolfqua_ropeca", "strategy_lm, Contrasts, ContrastsModerated, ContrastsROPECA", "peptide", "peptide.txt" )
)

>>>>

*R*'s formula interface for linear models is flexible, widely used, and well documented [@faraway2016extending; @law2020guide]. We use the formula interface to specify the models, making it easy to reproduce an analysis performed with *prolfqua* in other statistical programming languages. In addition, we implement features specific to high throughput experiments, such as the empirical Bayes variance and $p$-value moderation, which utilizes the parallel structure of the protein measurements and the analysis [@Ritchie2015]. We also compute probabilities of differential protein regulation based on peptide level models [@Suomi2017bEnhanced]. We used R6 classes to encapsulate the statistical modelling functionality in *prolfqua* (see Figure \@ref(fig:ContrastUML)). We specify a contrast interface (`ContrastsInterface`). Several implementations allow to perform DEA given linear or mixed effect models (`Contrasts`), variance shrinkage (`ContrastsModerated`), or to impute contrasts in cases when observations are missing for an entire group of samples (`ContrastsMissing`). Further implementations of the interface encapsulate and integrate DEA results of external tools such as *proDA* or of *SAINTexpress* [@teo2014saintexpress] used to analyze data from protein interaction studies.

@@

```{r prolfquaModels}
xM <- data.frame( 
    lm = c("strategy_lm, Contrasts", "linear modelling of peptide or protein abundances and group difference estimation"),
    lmer = c("strategy_lmer, Contrasts", "mixed effect modelling of peptide or protein abundances and group differences estimation" ),
    gmi = c("ContrastsMissing" , "group difference estimation when no observations in one of the groups" ),
    ROPECA = c("ContrastsROPECA", "estimating group differences for proteins by summarizing peptide differences"),
    limma = c( "ContrastsModerated", "empirical Bayes variance shrinkage for group difference estimates (limma)"),
    pic = c("runSaint, ContrastsSaintExpress", "protein interaction scoring (SAINTExpress)"),
    proDA = c( "strategy_proDA*, ContrastsProDA", "adapter to the probabilistic dropout model implemented in proDA"), check.names =  FALSE)
xM <- data.frame(t(xM))

@@

The following code example illustrates how we compute differences among groups. First, the linear model and the differences are specified. Afterward, the model is fitted to the data using the `build_model` function. Next, we estimate the contrasts either from the linear model using the `Contrasts` class or directly from the data using the `ContrastsMissing` class. Afterward, we apply $t$-statistic moderation using the `ContrastModerated` class. Finally, the `merge_contrasts_results` function merges both sets of contrast estimates, preferring the one obtained from the linear model if both are available. Then we create the plots shown in Figure \@ref(fig:figure2). Panel A shows the distribution of the $p$-values, Panel B shows the volcano plot for each comparison, and Panel C shows a Bland-Altman plot reporting the difference between groups as a function of the rank of the protein abundance.

@@

## estimate contrasts when observations missing
conI <- prolfqua::ContrastsMissing$new( lfqpro, contrasts, modelName = "mis")

@@

allBenchmarks <- allBenchmarks[c("benchmark_missing","benchmark_ProtModerated",  "benchmark_mixedModerated", "benchmark_ropeca","benchmark_merged","benchmark_mssstats","benchmark_proDA", "benchmark_msqrob")]

@@

```{r describeBenchmarkedModels, results='asis' }
xMM <- data.frame(
    msstats = c("MSstats", "preprocess with default parameters",  "", "evidence.txt"),
    msqrob = c("msqrob2", "hurdle msqrob model", "protein and peptide", "peptide.txt"),
    proDA = c("proDA", "probabilistic dropout model",  "protein", "peptide.txt"),
    imputation = c("prolfqua_missing", "ContrastsMissing, ContrastsModerated", "protein",  "peptide.txt"),
    lmmed = c("prolfqua_lm_mod", "strategy_lm, Contrasts, ContrastsModerated",  "protein", "peptide.txt"),
    prot_merged = c("prolfqua_merged", "addContrastResults( prefer = proflqua_lm_mod , add = prolfqua_missing)HACKIT ",  "protein", "peptide.txt"),
    mixed = c("prolfqua_mix_eff_mod", "strategy_lmer, Contrasts, ContrastsModerated", "peptide", "peptide.txt" ),
    ropeca = c("prolfqua_ropeca", "strategy_lm, Contrasts, ContrastsModerated, ContrastsROPECA", "peptide", "peptide.txt" )
)
